---
title: "Clean spectral data"
author: "Victoria Scholl"
date: "5/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
```

## Read original spectra

Spectra extracted using NEON data, by mbjoseph: https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d 

Columns: 

* **individualID** - unique identifier for each individual plant. Individuals can have multiple measurements over time.
* **date.x** - date of plant measurement/observation
* **uid.x** - unique identifier for the record
* **namedLocation** - name of measurement location in NEON database
* **adjEasting, adjNorthing** - UTM coordinates of mapped stem 
* **plantStatus** - physical status of the individual (live, dead, lost)
* **taxonID** - species code
* **scientificName** - scientific name, associated with taxonID, lowest taxonomic rank that can be determined.
* **band_idx** - number of the spectral band (band1, band 2, ... band 426)
* **wavelength_nm** - band wavelength in units of nm
* **reflectance** - decimal reflectance [0,1] extracted from NEON NIS hyperspectral imagery collected during same year
* **mask** - TRUE/LFALSE to identify water absorption bands


```{r}
all_spectra <- read.csv(here::here("data","all_spectra.csv"))
#colnames(all_spectra)
#summary(all_spectra)
```




